---
title: "Visualizing Distributions"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Liz Carey, Meghana Murthy"
date: "5/14/2019"
output:  
    html_document: 
      toc: true
      toc_float: true
      theme: paper
      code_folding: show
      highlight: textmate
      df_print: paged
    
---

## Set up  

Before we begin, install and load necessary packages to run this notebook. 

```{r, include = FALSE}
# install.packages("tidyverse")
# install.packages("gapminder")
```


```{r, warning = FALSE, message=FALSE} 

library(tidyverse) #includes functions to run code in this notebook and
library(gapminder) #includes dataset we'll be plotting
```

## What is a histogram? 

A histogram is a visualization of the distribution of numerical data. It's created by "binning" or bucketing the numerical values into intervals. 

### When to use a histogram:   
+ Check the distribution (shape) of your data 
+ Check the data for outliers

### How is a histogram different from a bar chart? 
+ A bar chart compares two variables while a histogram only relates one. 
+ A bar chart plots categorical data while a histogram plots numerical binned data
+ A bar chart can be reordered while a histogram cannot
+ A bar chart uses the length of a bar to display value while a histogram uses the area of the bar

## Histogram shapes  
There are various shapes that your histogram can appear

### Normal Distribution
A bell-shaped curve.  

```{r, warning = FALSE, message=FALSE, echo = FALSE, fig.align="center"}
norm_dist <-as.tibble(rnorm(1000, 0, 1))

ggplot(norm_dist, aes(x = value)) + 
  geom_histogram()

```


### Left skew
The data are clustered around a large value.  

```{r, warning = FALSE, message=FALSE, echo = FALSE, fig.align="center"}
left_skew <- as.tibble(rnorm(1000, mean=2, abs(log(rnorm(1000) ))))

  ggplot(left_skew, aes(x = value)) + 
  geom_histogram()


```


### Right skew  
The data are clustered around a small value 

```{r}
right_skew <- as.tibble(rnbinom(1000, mu = 10, size = 1))
  
  ggplot(right_skew, aes(x = value)) + 
  geom_histogram()
```


### Comb distribution
The distribution alternates between high and low peaks. A comb shape can be caused by rounding off. For example, if you are measuring water height to the nearest 10 cm and your class width for the histogram is 5 cm, this could cause a comb shape.


## Plotting a histogram



#### Print the data
By printing the data, we get a feel for what we'll be analyzing. 
From the table, we see there are the life expectancy, population, and GDP per capita of various countries separated by year. 

```{r, warning = FALSE, message=FALSE}
gapminder
```

```{r}
ggplot(gapminder, aes(lifeExp)) + 
  geom_histogram()
```

### A basic histogram
```{r, warning = FALSE}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram()
```

### Adjusting the bin width  

Adjusting the binwdith of your data can change the interpretation.

```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(bins = 200)
```


### Changing the colors of your histogram

```{r, include = FALSE}

#Store laCroix color palette, taken from the LaCroisColoR package
PassionFruit <- c("#C70E7B", "#FC6882", "#A6E000", "#1BB6AF", "#6C6C9D", "#172869")
Mango <- c("#FF5300", "#9ED80B", "#43B629", "#1BB6AF", "#8F92A1", "#172869")
Pure <- c("#AFDFEF", "#54BCD1", "#1BB6AF", "#0099D5", "#007BC3", "#172869")
Lime <- c("#2CB11B", "#95C65C", "#BDDE9B", "#1BB6AF", "#0076C0", "#172869")
Lemon <- c("#F7AA14", "#F5D000", "#F7E690", "#1BB6AF", "#088BBE", "#172869")
Orange <- c("#EF7C12", "#FCA315", "#F4B95A", "#1BB6AF", "#088BBE", "#172869")
Berry <- c("#B25D91", "#CB87B4", "#EFC7E6", "#1BB6AF", "#088BBE", "#172869")
CranRaspberry <- c("#D9565C", "#F28A8A", "#EDA9AB", "#1BB6AF", "#088BBE", "#172869")
Pamplemousse <- c("#EA7580", "#F6A1A5", "#F8CD9C", "#1BB6AF", "#088BBE", "#172869")
PeachPear <- c("#FF3200", "#E9A17C", "#E9E4A6", "#1BB6AF", "#0076BB", "#172869")
Coconut <- c("#881C00", "#AF6125", "#F4E3C7", "#1BB6AF", "#0076BB", "#172869")
Apricot <- c("#D72000", "#EE6100", "#FFAD0A", "#1BB6AF", "#9093A2", "#132157")
Tangerine <- c("#EF562A", "#EC921D", "#F7B449", "#FFED00", "#1BB6AF", "#9093A2", "#132157")
KeyLime <- c("#D84D16", "#FFF800", "#8FDA04", "#009F3F", "#132157")
PommeBaya <- c("#C23A4B", "#FBBB48", "#EFEF46", "#31D64D", "#132157")
CeriseLimon <- c("#EE4244", "#F8D961", "#B6D944", "#638E6E", "#3C5541", "#132157")
PinaFraise <- c("#F44B4B", "#F19743", "#F1F1A8", "#92D84F", "#7473A6", "#132157")
KiwiSandia <- c("#D18F55", "#FF3F38", "#FF8C8D", "#AFDE62", "#3CBC38", "#4F5791", " #132157")
MelonPomelo <- c("#EE404E", "#F99D53", "#F9E7C2", "#24C852", "#4F5791", "#132157")
MurePepino <- c("#D64358", "#EAFB88", "#3C8C4D", "#DFCEE0", "#4F5791", "#132157")
```

```{r, warning = FALSE, message=FALSE}
#store continents in a list
continents <- c("Africa", "Americas", "Asia", "Europe", "Oceania")

#create a function that will assign the continent names to a color. The function takes three inputs: the list of variables to be colored (e.g. continent names), the type of laCroix palette, and the start of the index. change the index_start to 0 or 1 to get variations of laCroix palette.
set_continent_color <- function(continent_names, lacroix_color, index_start) {
  
  #create empty vector to store values
  output <- vector("character", length(continent_names))
  
  #loop over the length of the list of continent names, and assign a color to each continent
  for (i in seq_along(continent_names)) {
    output[[i]] <- lacroix_color[[index_start + i]]
  }
  output 
  
}

```

#### Plot the histogram.  

Set custom colors using the `set_continent_color` function in the `scale_fill_manual` function.

<div class = "row">
  
<div class = "col-md-6">
```{r, warning = FALSE, message=FALSE, fig.align="center"}

ggplot(gapminder, aes(lifeExp, fill = continent)) +
  geom_histogram() + 
  scale_fill_manual(values = set_continent_color(continents, PeachPear, 1))
```
</div>

<div class = "col-md-6">
```{r, warning = FALSE, message=FALSE, fig.align="center"}

ggplot(gapminder, aes(lifeExp, fill = continent)) +
  geom_histogram() + 
  scale_fill_manual(values = set_continent_color(continents, Lemon, 1))
```  
</div>

</div>
  
#### Separate plots  

```{r}
ggplot(gapminder, aes(lifeExp, fill = continent)) +
  geom_histogram() + 
  scale_fill_manual(values = set_continent_color(continents, Pamplemousse, 1)) +
  facet_wrap(~continent)
```



## Density Plots  
A density plot is similar to a histogram, but instead of 

```{r, warning = FALSE}
ggplot(gapminder, aes(lifeExp, stat(density), colour = continent)) +
  geom_freqpoly()
```
```{r, warning=FALSE}
# Build dataset with different distributions
data <- data.frame(
  type = c( rep("Edge Peak", 1000), 
            rep("Comb", 1000), 
            rep("Normal", 1000), 
            rep("Uniform", 1000), 
            rep("Bimodal", 1000), 
            rep("Skewed", 1000) ),
  value = c( rnorm(900), rep(3, 100), 
             rnorm(360, sd=0.5), rep(c(-1,-0.75,-0.5,-0.25,0,0.25,0.5,0.75), 80), 
             rnorm(1000), runif(1000), 
             rnorm(500, mean=-2), 
             rnorm(500, mean=2), abs(log(rnorm(1000))) )
)

# Represent it
data %>%
  ggplot( aes(x=value)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.9, adjust = 0.5) +
    facet_wrap(~type, scale="free") +
    theme(
      panel.spacing = unit(0.1, "lines"),
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank()
    )
```

## Common pitfalls  


*** 

## Resources

Read more about different visualization types and how to build them from the following resources:  

+  [Data to Viz](https://www.data-to-viz.com/)  
    *From Data to Viz leads you to the most appropriate graph for your data. It links to the code to build it and lists common caveats you should avoid.*

